---
title: "Technical File: Food Swamps, Food Deserts, and Food Oases"
csl: apa.csl
output:
  word_document:
    toc: false
    reference_docx: word-styles-reference.docx
  pdf_document:
    toc: false
    toc_depth: 2
    number_section: true 
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: flatly
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo    = knitr::is_html_output(),  # code only in HTML
  results = knitr::is_html_output(),  # results only in HTML  
  warning = FALSE,
  message = FALSE
)

library(here)
here::i_am("technical-file-food-env.Rmd")
library(tidyverse)
library(sf)
library(tidycensus)
library(knitr)
library(ggplot2)
library(forcats)
library(purrr)
library(janitor)
options(tigris_class = "sf", tigris_use_cache = TRUE)
```


# Overview

This study examines how living in a food swamp, food desert, or food oasis shapes community health outcomes, with a focus on Virginia Beach City County and Henrico County. Using spatial analysis and Census data, we investigate how food environments relate to socioeconomic conditions and public health disparities.

# Literature Review

Prior research shows that limited access to supermarkets increases reliance on energy-dense, processed foods, contributing to obesity, hypertension, and diabetes. Structural barriers such as income constraints and transportation limitations further restrict access to healthy food options. Federal policies have attempted but disparities persist in low-income and predominantly Black communities (Walker, Keane, & Burke, 2010).

# Conceptual Framework

*Food Desert*: A geographic area where there is limited access to full-service supermarkets and healthy food options.

*Food Swamp*: A geographic area where convenience stores, fast food restaurants, restaurants, or liquor stores, generally unhealthy food options, heavily outnumber healthier alternatives.

*Food Oasis*: A geographic area where food outlets and grocery stores that offer healthy food options, indicate relatively strong access to nutritious food sources within the local food environment.

# Research Questions

How are tract-level food environments associated with socioeconomic conditions in VA Beach City and Henrico Counties?

# Methods and Data

Data were drawn from the Department of Agriculture Economic Research through the Food Access Research Atlas. The Atlas defines low-access tracts based on distance to the nearest supermarket, while also incorporating income to identify vulnerable populations and construct classifications. We integrate U.S public health data and the `tidycensus()` package in R to conduct an initial case study of food environments in Henrico and Virginia Beach City Counties.

```{r, eval=F, include=F, warning=F, message=F}
# Load food access data
food.access <- read.csv("data/food-access/foodaccessdata2019.csv")

# Filter Virginia
va_food <- food.access %>%
  filter(State == "Virginia") %>%
  mutate(GEOID = sprintf("%011.0f", CensusTract))

# Census data
va_census <- get_acs(
  geography = "tract",
  state = "VA",
  variables = c(
    total_pop = "B01003_001",
    median_income = "B19013_001"
  ),
  year = 2020,
  survey = "acs5"
)

va_census_wide <- va_census %>%
  select(GEOID, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate)

# Merge
va_merged <- va_food %>%
  left_join(va_census_wide, by = "GEOID")

saveRDS(va_merged, file = "va_merged.rds")

```


```{r, include=F, warning=F, message=F}
va_merged <- readRDS("data/va_merged.rds")

# Clean
va_clean <- va_merged %>%
  mutate(
    across(where(is.character), ~na_if(., "NULL")),
    across(
      c(PovertyRate, MedianFamilyIncome, lalowihalfshare,
        lasnaphalfshare, lapophalfshare),
      readr::parse_number
    )
  )

# Study counties
va_study <- va_clean %>%
  filter(County %in% c("Virginia Beach city", "Henrico County"))

# Classifications
snap_cut   <- median(va_study$lasnaphalfshare, na.rm = TRUE)
access_cut <- median(va_study$lapophalfshare, na.rm = TRUE)

va_study <- va_study %>%
  mutate(
    food_desert = LILATracts_1And10 == 1,
    food_oasis  = LILATracts_1And10 == 0 & LowIncomeTracts == 0,
    food_swamp  = LowIncomeTracts == 1 &
                  lasnaphalfshare > snap_cut &
                  lapophalfshare > access_cut,
    food_env = case_when(
      food_desert ~ "Desert",
      food_swamp  ~ "Swamp",
      food_oasis  ~ "Oasis",
      TRUE        ~ "Other"
    )
  )

# Spatial data
va_tracts <- get_acs(
  geography = "tract",
  state = "VA",
  county = c("Virginia Beach city", "Henrico County"),
  variables = "B01003_001",
  geometry = TRUE
)

va_map <- va_tracts %>%
  select(GEOID, geometry) %>%
  left_join(va_study, by = "GEOID")

# Split maps
va_beach_map <- va_map %>%
  filter(County == "Virginia Beach city")

henrico_map <- va_map %>%
  filter(County == "Henrico County")

food_levels <- c("Desert", "Swamp", "Oasis")
food_colors <- c(
  "Desert" = "red",
  "Swamp"  = "orange",
  "Oasis"  = "green"
)

va_map <- va_tracts %>%
  select(GEOID, geometry) %>%
  left_join(va_study, by = "GEOID") %>%
  mutate(
    food_env = factor(food_env, levels = food_levels)
  )

map_va_beach <- ggplot(filter(va_map, County == "Virginia Beach city")) +
  geom_sf(aes(fill = food_env), color = NA, show.legend = TRUE) +
  scale_fill_manual(
    values = food_colors,
    drop = FALSE,
    limits = food_levels,
    na.value = "grey85",
    name = "Type"
  ) +
  labs(title = "Virginia Beach City County") +
  theme_minimal() +
  theme(legend.position = "bottom")

map_henrico <- ggplot(filter(va_map, County == "Henrico County")) +
  geom_sf(aes(fill = food_env), color = NA, show.legend = TRUE) +
  scale_fill_manual(
    values = food_colors,
    drop = FALSE,
    limits = food_levels,
    na.value = "grey85",
    name = "Type"
  ) +
  labs(title = "Henrico County") +
  theme_minimal() +
  theme(legend.position = "bottom")

```

# Analytic Method

For our study, 'Food Desert' denotes USDA low-income, low-access (LILA) tracts, "under the 1-mile urban / 10-mile rural Food Access Research Atlas measure", 'Food Oasis' denotes tracts that are neither low income, nor low access. 'Food Swamp' is a study-specific proxy based on tract-level, "access and SNAP-related measures" (U.S. Dept of Agriculture Economic Research, 2019).

# Findings

```{r, message=F, warning=F}
map_va_beach
```

```{r, warning=F, messages=F}
map_henrico
```

Grey (Other) includes tracts not meeting any study definition.

# Discussion

The majority of the VA Beach area shows up as a food oasis. However, access to and use of these food resources is not equitable.

Similarly, in Henrico County, red areas (deserts) correlate with limited access (spatially isolated) and social demographics (i.e., predominately Black and lower income) by neighborhood.

Findings suggest that lower-income areas are still likely to be classified as food swamps or food deserts even in tourist areas. Tourism matters because it shows up indirectly through land use, food retail incentives, and who food systems are built to serve. Tourist-heavy corridors can also distort how "access" looks on paper (Block, Scribner, & DeSalvo, 2004).

# Conclusion

Food environments play a critical role in shaping community health. Policies aimed at improving access to healthy food—such as incentivizing grocery store development—may help reduce disparities in diet-related health outcomes. However, more work needs to be done on citizen access to food in areas where tourism may mask the presence of food deserts or food swamps.

```{r, message=F, warning=F}
# install packages
#install.packages("tidycensus")
#install.packages("tidyverse")
#install.packages("mapview")
#install.packages("mapgl")
#install.packages("sf") 
#install.packages("viridis")
library(tidycensus)
library(dplyr)
library(ggplot2)
library(sf)
library(viridis)
library(mapview)
```

# virginia beach
Gather spatial data for Virginia Beach, Virginia  

```{r, messages=FALSE, warning=FALSE}
va_income_sf = get_acs(
  geography = "county",
  variables = "B19013_001", # income variable
  state = "VA",
  year = 2023,
  geometry = T
  )


plot(va_income_sf['estimate'])

```

```{r, messages=FALSE, warning=FALSE}

va_henrico_income_sf = get_acs(
  geography = "tract",
  county = "Henrico",
  variables = "B19013_001", # income variable
  state = "VA",
  year = 2023,
  geometry = T
)

```

# henrico

```{r, messages=FALSE, warning=FALSE}
ggplot(va_henrico_income_sf) +
  geom_sf(aes(fill = estimate), color = "white", size = 0.1) +
  scale_fill_viridis(option = "magma", name = "Median Household Income") +
  labs(
    title = "Median Household Income in Henrico County, VA",
    subtitle = "By Census Tract (2023)",
    caption = "Data: American Community Survey 5-Year Estimates"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.title = element_blank(),
    legend.position = "right"
  )
```

I am about to make map of Richmond city.


```{r, message=F, warning=F}

# install packages
#install.packages("tidycensus")
#install.packages("tidyverse")
#install.packages("mapview")
#install.packages("mapgl")
#install.packages("sf") 
#install.packages("viridis")
library(tidycensus)
library(dplyr)
library(ggplot2)
library(sf)
library(viridis)
library(mapview)
library(tidycensus)
library(tidyverse)
library(sf)
library(viridis)
library(mapview)
library(gridExtra)
library(scales)
```

```{r, messages=FALSE, warning=FALSE}
va_counties <- get_acs(
  geography = "county",
  variables = c(black = "B02001_003", # Black/African American population alone
                total = "B01001_001" # Total population
  ),
  state = "VA",
  year = 2023,
  geometry = T,
  output = "wide"
)

# Add county FIPS code by extracting first 5 digits of GEOID
va_counties <- va_counties %>%
  mutate(county_fips = substr(GEOID, 1, 5)) %>% 
  relocate(GEOID, county_fips)

#View(va_counties)
```

```{r, messages=FALSE, warning=FALSE}
head(va_counties) %>% 
  select(GEOID, county_fips, NAME, blackE, totalE)
```

```{r, messages=FALSE, warning=FALSE}
va_tract <- get_acs(
geography = "tract",
variables = c(black = "B02001_003", # Black/African American population alone
population = "B01001_001", # Total population
immigrant = "B05002_014", # Immigrant population
education = "B06009_005", # Individuals that have a bachelor's degree
poverty = "B17001_001", # Individuals whose household income
# is below the poverty line
income = "B19001_001", # Household income
young1 = "B01001_007", # Male 18-19
young2 = "B01001_008", # Male 20
young3 = "B01001_009", # Male 21
young4 = "B01001_010" # Male 22-24
),
state = "VA",
year = 2023,
geometry = T,
output = "wide"
)
```

```{r, messages=FALSE, warning=FALSE}
va_counties %>% 
  arrange(desc(blackE)) %>%  # desc is descending order
  head(n=10)
```

```{r, messages=FALSE, warning=FALSE}
va_counties %>% 
  mutate(prop_black = blackE / totalE) %>% 
  relocate(GEOID, prop_black) %>% 
  arrange(desc(prop_black))
```

```{r, messages=FALSE, warning=FALSE}
va_counties %>% 
  mutate(prop_black = blackE / totalE) %>% 
  relocate(GEOID, prop_black) %>% 
  arrange(-desc(prop_black))
```

```{r, messages=FALSE, warning=FALSE}
virginia_beach_city_tracts <- get_acs(
  geography = "tract",
  variables = c(
    black = "B02001_003",      # Black or African American alone
    total = "B01001_001"       # Total population
  ),
  state = "VA",
  county = "Virginia Beach City",
  year = 2023,
  geometry = TRUE,
  output = "wide"
)

virginia_beach_city_tracts %>% 
  head()
```

```{r, messages=FALSE, warning=FALSE}
ggplot(virginia_beach_city_tracts) +
  geom_sf(aes(fill = blackE), color = NA) +
  scale_fill_viridis_c(option = "magma", na.value = "grey50", labels = comma) +
  labs(title = "Estimated Black Population by Census Tract in Virginia Beach City County, VA (2023)",
       fill = "Population") +
  theme_minimal()
```

```{r, messages=FALSE, warning=FALSE}
virginia_beach_city_tracts <- virginia_beach_city_tracts %>% 
  mutate(nonblackE = totalE - blackE) %>% 
  select(GEOID, blackE, nonblackE, totalE)
```

```{r, messages=FALSE, warning=FALSE}
ggplot(virginia_beach_city_tracts) +
  geom_sf(aes(fill = nonblackE)) +
  scale_fill_viridis_c(option = "magma", 
                       na.value = "grey50",
                       labels = comma) +
  labs(title = "Estimated Non-Black Population by Census Tract in Virginia Beach City (2023)",
       fill = "Population") +
  theme_minimal()
```

```{r, messages=FALSE, warning=FALSE}
virginia_beach_city_tracts <- virginia_beach_city_tracts %>% 
  mutate(proportion_black = blackE / totalE) %>%
  mutate(proportion_non_black = 1 - proportion_black)
```

```{r, messages=FALSE, warning=FALSE}
virginia_beach_city_tracts %>%
  arrange(desc(proportion_black)) %>% 
  select(GEOID, proportion_black, proportion_non_black) %>% 
  head(n = 10)
```

```{r, messages=FALSE, warning=FALSE}
virginia_beach_city_tracts %>%
  st_set_geometry(NULL) %>%   # Remove geometry to treat as normal data.frame
  summarise(
    total_tracts = n(),
    tracts_above_threshold = sum(proportion_black >= 0.60, na.rm = TRUE),
    proportion_above_threshold = mean(proportion_black >= 0.60, na.rm = TRUE)
  )
```

```{r, messages=FALSE, warning=FALSE}
virginia_beach_city_total_black <- sum(virginia_beach_city_tracts$blackE, na.rm = TRUE)
virginia_beach_city_total_nonblack <- sum(virginia_beach_city_tracts$nonblackE, na.rm = TRUE)

dissimilarity_virginia_beach_city <- 0.5 * sum(abs(
  (virginia_beach_city_tracts$blackE / virginia_beach_city_total_black) - 
  (virginia_beach_city_tracts$nonblackE /virginia_beach_city_total_nonblack)
), na.rm = TRUE)
dissimilarity_virginia_beach_city
```

```{r, messages=FALSE, warning=FALSE}
ggplot(virginia_beach_city_tracts) +
  geom_sf(aes(fill = proportion_black), color = "white") +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "Proportion of Black Residents by Census Tract in Virginia Beach City",
       fill = "Proportion Black") +
  theme_minimal()
```



```{r, messages=FALSE, warning=FALSE}
virginia_beach_city_tracts <- virginia_beach_city_tracts %>%
  mutate(majority_black = proportion_black > 0.5)
```

```{r, messages=FALSE, warning=FALSE}
# Union geometries by group
virginia_beach_city_union_black <- virginia_beach_city_tracts %>%
  filter(majority_black) %>%
  summarise(geometry = st_union(geometry))

virginia_beach_city_union_nonblack <- virginia_beach_city_tracts %>%
  filter(!majority_black) %>%
  summarise(geometry = st_union(geometry))
```

```{r, messages=FALSE, warning=FALSE}
virginia_beach_city_boundary_line = 
  st_intersection(st_boundary(virginia_beach_city_union_black), st_boundary(virginia_beach_city_union_nonblack))
```

```{r, messages=FALSE, warning=FALSE}
ggplot() +
  geom_sf(data = virginia_beach_city_tracts, aes(fill = majority_black), 
          color = "grey40", 
          alpha = 0.5) +
  geom_sf(data = virginia_beach_city_boundary_line, 
          color = "red", 
          size = 1) +
  labs(title = "Dividing Walls Between Majority Black and Non-Black Areas in Virginia Beach City, VA") +
  theme_minimal()
```